---
title: "Lab02: Logistic Regression"
subtitle: "Princeton University"
author: "James Packman"
output: 
  tufte::tufte_html:
    css: 
    tufte_variant: "envisioned"
    highlight: github-dark
    fig_height: 10
    fig_width: 16
    toc: true
    toc_depth: 1
execute: 
  message: false
  warning: false
format: 
  html:
    code-fold: true
    code-overflow: wrap
engine: knitr
---

-   Assignment requirements:

    -   If you are using Github (recommended), make sure to commit and push your work to GitHub regularly, at least after each exercise. Write short and informative commit messages, and share the link to your assignment with me. If not, you can also send me the **rmd** & **rendered file** via Canvas.

    -   In this assignment, you will not need to code from scratch. Rather, you'll need to fill in code where needed. This assignment has a logisitic regression implementation for a scenario from EDA down to model comparison (and would be useful for whenever you may encounter such a situation in the future).

    -   I want the assignments to begin reflecting a bit more of how you'd be doing things on your own, where you have some prior knowledge and you figure other things out (by referring to documentation, etc.) . In addition to the rmd, I also want you to submit to me **notes** of anything new that you learn while finishing the assignment. And any pain-points, and we'll discuss more.

-   Note:

    -   If you are fitting a model, display the model output in a neatly formatted table. (The `gt` `tidy` and `kable` functions can help!). `Modelsummary` also looks good(inst)

    -   Make sure that your plots are clearly labeled – for all axes, titles, etc.\

## Data: General Social Survey

The General Social Survey (GSS) has been used to measure trends in attitudes and behaviors in American society since 1972. In addition to collecting demographic information, the survey includes questions used to gauge attitudes about government spending priorities, confidence in institutions, lifestyle, and many other topics. A full description of the survey may be found here.

The data for this lab are from the 2016 General Social Survey. The original data set contains 2867 observations and 935 variables. We will use and abbreviated data set that includes the following variables:

natmass: Respondent’s answer to the following prompt:

“We are faced with many problems in this country, none of which can be solved easily or inexpensively. I’m going to name some of these problems, and for each one I’d like you to tell me whether you think we’re spending too much money on it, too little money, or about the right amount…are we spending too much, too little, or about the right amount on mass transportation?”

age: Age in years.

sex: Sex recorded as male or female

sei10: Socioeconomic index from 0 to 100

region: Region where interview took place

polviews: Respondent’s answer to the following prompt:

“We hear a lot of talk these days about liberals and conservatives. I’m going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal - point 1 - to extremely conservative - point 7. Where would you place yourself on this scale?”

The data are in gss2016.csv in the data folder.

## EDA

> Let’s begin by making a binary variable for respondents’ views on spending on mass transportation. Create a new variable that is equal to “1” if a respondent said spending on mass transportation is about right and “0” otherwise. Then plot the proportion of the response variable, using informative labels for each category.

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(modelsummary)
library(tidyr)
library(knitr)
library(easystats)
library(broom)
library(emmeans)
library(marginaleffects)
library(performance)
library(arm)
library(modelsummary)
```

```{r}
# load data
data <- read.csv("gss2016.csv")

```

Fill in the "\_\_\_\_" below to encode the binary variable

```{r}
data <- data %>%
   mutate(mass_trans_spend_right = ifelse(data$natmass == "About right", 1, 0))
```

Note: Slight difficulty arose when I was unsure how to set the conditional for a character string. Remember to use == and "".

```{r}
#Get proportions
mass_spend_summary <- data %>%
  count(mass_trans_spend_right) %>%
  mutate(proportion = n / sum(n))

#Look at the dataframe structure. And make sure it's in a format that you can use for plotting.
#Change structure if needed
mass_spend_long <- pivot_longer(mass_spend_summary, names_to = "opinion", values_to = "proportion", cols = proportion)

#Factorise for plot
mass_spend_long$mass_trans_spend_right <- as.factor(mass_spend_long$mass_trans_spend_right)


#Make plot
#Hint: geom_bar lets you make stacked bar charts
ggplot(data = mass_spend_long, mapping = aes(x= "", y = proportion, fill = factor(mass_trans_spend_right))) + geom_bar(stat = "identity", show.legend = FALSE) + labs(title = "Respondents' Views of Mass Transportation Spending", subtitle = "2016 General Social Survey", x = "Spending is...", y = "Proportion") + annotate("text", x = "", y = 0.75, label = "Too Much or Too Little: 47%") +  annotate("text", x = "", y = 0.25, label = "About Right: 52.9%")

```

> Recode `polviews` so it is a factor with levels that are in an order that is consistent with question on the survey. Note how the categories are spelled in the data.

```{r}
data <- data %>%
  mutate(polviews = factor(polviews,
                           levels = c("Extremely liberal", "Liberal", "Slightly liberal", "Moderate", "Slghtly conservative", "Conservative", "Extrmly conservative"),
                           ordered = TRUE))
```

> Make a plot of the distribution of `polviews`

```{r}
#Get proportions, format, and produce a plot like you did previously for mass_trans_spend_right

#Get proportions
polviews_summary <- data %>%
  count(polviews) %>%
  mutate(proportion = n / sum(n))
  
#Format
polviews_long <- pivot_longer(polviews_summary, names_to = "political views", values_to = "proportion", cols = proportion)

#Factorise for plot
polviews_long$polviews <- as.factor(polviews_long$polviews)

#Make plot
ggplot(data = polviews_long, mapping = aes(x= "", y = proportion, fill = factor(polviews))) + geom_bar(stat = "identity", show.legend = TRUE, alpha = 0.6) + labs(title = "Respondents' Political Views", subtitle = "2016 General Social Survey") + geom_text(aes(label = round(proportion, digits = 3)), size = 3, hjust = 0.5, position = position_stack(vjust = 0.5))
                                                                                                                                         
```

> Which political view occurs most frequently in this data set?
>
> "Moderate" is the political view which occurs most frequently in this data set (about 37.6% of respondents identified as politically moderate).
>
> \_\_\_\_\_

> Make a plot displaying the relationship between satisfaction with mass transportation spending and political views. Use the plot to describe the relationship the two variables.

```{r}
#________
ggplot(data = data, mapping = aes(x = polviews, fill = factor(mass_trans_spend_right))) +
  geom_bar(position = "fill", show.legend = TRUE) +
  labs(x = "Political View",
       y = "Satisfaction with Transportation Spending",
       title = "Satisfaction with Spending vs. Political Views") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

The more conservative one's political views are the more they think the amount of spending on mass transportation is correct.

*Note*: I tried to label the bars directly instead of using a legend, but I could not figure out how to do so in an aesthetically pleasing manner.

-   We’d like to use age as a quantitative variable in your model; however, it is currently a character data type because some observations are coded as "89 or older".

> Recode age so that is a numeric variable. Note: Before making the variable numeric, you will need to replace the values "89 or older" with a single value.cc

```{r}

data <- data %>%
  mutate(age = if_else(age == "89 or older", '89', age), 
         age = as.numeric(age))
```

> Plot the frequency distribution of age.

```{r}
ggplot(data = data, aes(x = age)) +
  geom_histogram(stat = "bin", position = "stack", binwidth = 10) +
  labs(x = "Age",
       y = "Frequency",
       title = "Distribution of Age")
```

## Logistic regression

> Let’s start by fitting a logistic regression model with just the intercept

```{r}
intercept_only_model <- glm(
  data = data,
  family = binomial(link = "logit"),
  mass_trans_spend_right ~ 1
  ) 

intercept_only_model %>% 
  tidy() %>%
  kable()

```

> Interpret the intercept in the context of the data. You can do this by converting the $\beta_0$ parameter out of the log-odds metric to the probability metric. Make sure to include the 95% confidence intervals. Then interpret the results in a sentence or two--what is the basic thing this probability tells us about?

```{r}
b0 <- coef(intercept_only_model) # get coef

b0_transformed <- exp(b0) / (1 + exp(b0)) # logistic transform

ci_lower = b0 - 1.96 * 0.0393685
ci_upper = b0 + 1.96 * 0.0393685

#transforming confidence intervals of coefficients into probabilities
p_lower = exp(ci_lower) / (1 + exp(ci_lower))
p_upper = exp(ci_upper) / (1 + exp(ci_upper))

b0_transformed

```

Interpretation: b0_transformed is approximately 0.53, meaning that the odds of an average person in the sample saying they are satisfied with spending on mass transportation is about 53%.

> Now let's fit a model using the demographic factors - `age`,`sex`, `sei10` - to predict the odds a person is satisfied with spending on mass transportation. Make any necessary adjustments to the variables so the intercept will have a meaningful interpretation. *Neatly display the model coefficients (do not display the summary output)*

```{r}
#make sure that sex is a factor (i.e. to make sure R knows it's binary/categorical, and not continuous)
data$sex <- relevel(factor(data$sex), ref="Male")


#fit with glm()
m1 <- glm(
  data = data,
  family = binomial(link = "logit"),
  mass_trans_spend_right ~ age + sex + sei10
  ) 

#produce tidy output of model coefficients
m1 %>% 
  tidy() %>%
  kable()
```

Note: For some ungodly reason, releveling the factor "sex" was incredibly difficult. Writing one line of code with as.factor and then a subsequent line releveling data\$sex produced an error telling me that relevel only works for unordered factors, but putting (factor) in the same line as the relevel worked.

> Consider the relationship between sex and one’s opinion about spending on mass transportation. Interpret the coefficient of sex in terms of the logs odds and OR of being satisfied with spending on mass transportation. What are the predicted probabilities for males and females on support for spending on mass transportation? Please include the 95% CIs around each estimate.

```{r}
m1 %>% 
  tidy() %>%
  kable()

m1 %>% 
  tidy(exponentiate = TRUE) %>%
  kable()

bsex <- coef(m1)["sexFemale"]
bsex

ci_lower_lo = bsex - 1.96 * 0.0798020
ci_upper_lo = bsex + 1.96 * 0.0798020

ci_lower_or = 1.29 - 1.96 * 0.0798020
ci_upper_or = 1.29 + 1.96 * 0.0798020

emm_sex <- emmeans(m1, "sex", type = "response")
emm_sex

#OR for comparison
pairs(emm_sex, reverse = TRUE)

```

If you did this right, you'll find that being female (as compared to male) is associated with an increase in the log-odds of being satisfied with spending on mass transportation by 0.2557439 units (95% CI \[0.09, 0.41\]), holding all other variables constant. This equates to the odds of thinking the spending amount is right in females being 1.29 times the odds of thinking this in men (95% CI \[1.13, 1.44\]).

The predicted probability for females to be satisfied with spending on mass transportation is 55.9% (95% CI \[53.3%, 58.5%\]) and that of males is 49.5% (95% CI \[46.7%, 52.4%\]).\
\
Verify this.

Yes, my output replicates the listed results.

> Next, consider the relationship between age and one’s opinion about spending on mass transportation. Interpret the coefficient of age in terms of the logs odds and OR of being satisfied with spending on mass transportation. Please include the 95% CIs around each estimate.

```{r}

m1 %>% 
  tidy() %>%
  kable()

m1 %>% 
  tidy(exponentiate = TRUE) %>%
  kable()

bage <- coef(m1)["age"]

age_ci_lower_lo = bage - 1.96 * 0.0022824
age_ci_upper_lo = bage + 1.96 * 0.0022824

age_ci_lower_or = 0.9938530 - 1.96 * 0.0022824
age_ci_upper_or = 0.9938530 + 1.96 * 0.0022824





```

A one unit increase in age is associated with a decrease in the log-odds of being satisfied with spending on mass transportation by 0.00616 units (95% CI \[-0.0106, -0.00169\]), holding all other variables constant. The odds ratio is less than 1 (0.994) which confirms the negative relationship implied by the log-odds coefficient. Specifically, for each additional unit of age, the odds of being satisfied with mass transportation spending decrease by a factor of about 0.994, or approximately 0.6% per unit increase in age, holding other factors constant (95% CI \[0.989, 0.998\]).

> Consider the relationship between SES and one’s opinion about spending on mass transportation. Interpret the coefficient of SES in terms of the logs odds and OR of being satisfied with spending on mass transportation. Please include the 95% CIs around each estimate. ß

```{r}
bses <- coef(m1)["sei10"]

sei10_ci_lower_lo = bses - 1.96 * 0.0016609
sei10_ci_upper_lo = bses + 1.96 * 0.0016609

sei10_ci_lower_or = 	0.9937922 - 1.96 * 0.0016609
sei10_ci_upper_or = 	0.9937922 + 1.96 * 0.0016609

```

A one unit increase in SES index is associated with a decrease in the log-odds of being satisfied with spending on mass transportation by 0.0062 units (95% CI \[-0.0948, -0.00297\]), holding all other variables constant. The odds ratio is less than 1 (0.9937922), which confirms the negative relationship implied by the log-odds coefficient. Specifically, for each additional unit of SES index, the odds of being satisfied with mass transportation spending decrease by a factor of about 0.993, or approximately 0.7% per unit increase in SES index, holding other factors constant (95% CI \[0.991, 0.997\]).

## Marginal effects

-   Let's examine the results on the probability scale.

> Calculate the marginal effects of sex, age, and SES on mass transportation spending. You can use the `margins` package function `margins` discussed in your textbook or you can use the `marginaleffects` package `avg_slope` `avg_comparisons` discussed in lecture. Interpret each estimate.

```{r}

#Marginal effects
library(margins)
marg1 <- margins(m1)
summary(marg1)

avg_slopes(m1)%>%kable()

avg_comparisons(m1, comparison = "difference") %>% 
  kable()

```

-   The marginal effect of age is -0.0015 (95% CI \[-0.0026, -0.0004\]). So, for each additional unit increase of age, the probability of being satisfied with mass transportation spending decreases by approximately 0.15 percentage points, holding other factors constant (p = 0.0066).

-   The marginal effect of SES is also -0.0015 (95% CI \[-0.0023, -0.0007\]). For each one-unit increase in the socioeconomic index, the probability of being satisfied with mass transportation spending decreases by approximately 0.15 percentage points, holding other variables constant.

-   The marginal effect for being female compared to male is 0.0631 (95% CI \[0.0246, 0.102\]). This indicates that females are, on average, about 6.3 percentage points more likely than males to be satisfied with mass transportation spending, holding other factors constant.

## Model comparison

-   Now let’s see whether a person’s political views has a significant impact on their odds of being satisfied with spending on mass transportation, after accounting for the demographic factors.

> Conduct a drop-in-deviance/likelihood ratio test to determine if polviews is a significant predictor of attitude towards spending on mass transportation. Name these two models `fit2` and `fit3`, respectively. Compare the two models.

```{r}

fit2 <- glm(data = data, 
            family = binomial,
            mass_trans_spend_right ~ age + sex + sei10)

fit3 <- glm(data = data, 
            family = binomial,
            mass_trans_spend_right ~ age + sex + sei10 + polviews)

test_likelihoodratio(fit2, fit3) %>% kable()


```

> Is the model with polviews better than the model without?

-   Yes, it appears that the model is better with polviews.

## Visualization

-   Let’s plot the results

-   We next use the model to produce visualizations:

    -   Given the code below, interpret what is being plotted:

        -   pol_plot : Predicted probability of reporting being satisfied with mass transportation spending for each political view.

        -   sex_plot : Predicted probability of reporting being satisfied with mass transportation spending for each gender group.

        -   ses_plot: The predicted probability of satisfaction changes vs SES, a continuous variable.

    ::: callout-tip
    -   adjust the various settings in your plot to make it look professional.

    -   You can use `ggeffects` to get the predicted probabilities for these models.
    :::

```{r}
#| fig-width: 12
#| fig-height: 10
library(ggeffects)


colors <- c("Extremely liberal" = "black",
            "Liberal" = "#0e2f44",  # Dark blue
            "Slightly liberal" = "#1d5a6c",  # Less dark blue
            "Moderate" = "#358ca3",  # Medium blue
            "Slghtly conservative" = "#71b9d1",  # Light blue
            "Conservative" = "#a6dcef",  # Lighter blue
            "Extrmly conservative" = "#d0f0fd")  # Very light blue

pp_pol <- ggemmeans(fit3, terms = c("polviews"))

# Adjusted plot with gradient colors
pol_plot <- ggplot(pp_pol, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 2, show.legend = FALSE) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  scale_color_manual(values = colors) +
  labs(title = "Effect of Political Views on Satisfaction with Mass Transportation",
       x = "Political Views", y = "Predicted Probability",
       color = "Political Views") +
  theme_minimal()

pol_plot

pp_sex <- ggemmeans(fit3, terms = c("sex"))

sex_plot <- ggplot(pp_sex, aes(x = x, y = predicted, color = x)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2, show.legend = FALSE) +
  labs(title = "Effect of Sex on Satisfaction with Mass Transportation",
       x = "Sex", y = "Predicted Probability",
       color = "Sex") +
  theme_minimal()

pp_sex
sex_plot

pp_ses <- ggemmeans(fit3, terms = "sei10 [all]")


ses_plot <-  ggplot(pp_ses, aes(x = x, y = predicted)) +
  geom_line(color = "#2c7fb8", size = 1) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#2c7fb8", alpha = 0.2) +  # Add a confidence interval band
  labs(title = "Effect of SES on Satisfaction with Mass Transportation",
       x = "Socioeconomic Status Index (higher numbers indicate higher status)", y = "Predicted Probability") +
  theme_minimal() +
  theme(legend.position = "none")  
ses_plot

```

Note: no matter in which layer I put "show.legend = FALSE," I can't get the legends to disappear.

Question: What is the difference between ggmeans and ggeffects for calculating the predicted probabilities?

## Model Assumptions

-   Is the logistic model a good choice for this data?

```{r}
library(easystats)
#fit2
binned_residuals(fit2)
check_model(fit2, check = c("binned_residuals", "outliers", "vif"), residual_type = "normal")

#fit3
binned_residuals(fit3)
check_model(fit3, check = c("binned_residuals", "outliers", "vif"), residual_type = "normal")
```

::: callout-note
Answer: Model assumptions are generally met. VIF is low for both fit2 and fit3, meaning we do not have to worry about multicolinearity. There also do not seem to be outliers. As for binned residuals, 86% of the residuals are within the error bounds for fit2, and 90% of residuals are in error bounds for fit3. A warning indicates that "\~95% or higher would be good," so fit2 and fit3 fall a little short on this criterion.
:::

## Model fit

-   Calculate the $R^2$ for this model

```{r}

r2_mcfadden(fit2)

r2_mcfadden(fit3)

```

> R2 interpretation: Fit3, which includes polviews, has a higher McFadden's Pseudo R2, which suggests improvement over fit2 in goodness of fit.
>
> Next, Take a look at the binned residual plots for each continuous predictor variable and look at linearity. Is there a predictor that sticks out? What can we do to improve model fit in this case?

```{r}
binned_residuals(fit2, term="sei10")
binned_residuals(fit2, term="age")

binned_residuals(fit2, term="sei10") %>% plot(show_dots=TRUE)
binned_residuals(fit2, term="age") %>% plot(show_dots=TRUE)

```

::: callout-note
-   ::: callout-note
    Sei10 has more residuals outside of the error bounds. Since some of the bins have average residuals with large magnitudes, that might suggest the need for an interaction term. For example, SES might interact with age or political views.
    :::
:::

## Testing Polviews

```{r}
emmeans(fit3, "polviews") %>% pairs() %>% as.data.frame() %>% filter(p.value < .05)

emmeans(fit3, "polviews", type="response") %>% pairs() %>% as.data.frame() %>% filter(p.value < .05)
```

-   Conservatives are 2.70 times more likely to support mass transit spending compared to extremely liberal people and 2.17 times more likely to support mass transit spending compared to liberals (Question: I took the inverse of the odds ratios above to calculate these values. Is this correct?)

-   Extreme liberals are 0.370, 0.396, and 0.428 times more likely to support spending compared to conservatives, moderates and slight conservatives, respectively (Question: this would just be the odds ratios as depicted in the output above, correct?)

-   Extreme conservatives are 3.07 times more likely to support mass spending than liberals and 2.08 times more likely than slight liberals.

-   Liberals are 0.492 times and 0.532 times more likely (i.e., about half as likely) to support spending than moderates and slight conservatives, respectively.

## Conclusion

\_\_\_ based on the model summary below, and the three figures.

Satisfaction with mass transit spending (dichotomous) was modeled as a function of age, gender (male, female), and SES. Female respondents and respondents identifying as more politically conservative demonstrated higher predicted probabilities of reporting satisfaction with transportation spending. SES and age, which were continuous variables, had a negative relationship with predicted probability of satisfaction, such that every unit increase in age and SES predicted decreases in the likelihood of reporting satisfaction. Does the model summary indicate that including each of the predictors significantly improves the model?

```{r model-summary, results='asis', echo=FALSE , fig.cap="Table 1"}

fit_anova = anova(fit3, test="Chisq")

fit_anova %>%
  kable()


```

```{r plot-sexviews, fig.cap="Figure 1: Effect of Sex on Satisfaction with Mass Transportation", echo=FALSE}
sex_plot
```

```{r plot-sesviews, fig.cap="Figure 2: Effect of SES  on Satisfaction with Mass Transportation", echo=FALSE}
ses_plot
```

```{r plot-polviews, fig.cap="Figure 3: Effect of Political Views on Satisfaction with Mass Transportation", echo=FALSE}
pol_plot
```
